---
title: "Practical Machine Learning - Course Project"
author: "Neal V. Quizon"
date: "December 9, 2018"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Practical Machine Learning - Course Project

#Introduction
The aim of the project is to predict how well an exercise (in our project specifically is barbell lifts) is done according to a set of variables that have been derived using sensors applied on the body.


The objective is to correctly predict the variable classe of the Test set. This variable indicates how well the exercise is performed. The valua A indicates that the exercise was well performed while the other letters (from B to E) respectively indicate that common mistakes has been done during the execution of the weightlifting.

First the datasets are loaded and only useful variables are considered. Then three different Machine learning algorithm are applied to a subset of the training set and then tested to estimate the accuracy. Finally, the best model found (i.e. Random forest) is applied to the test set to predict the type of performance in doing the weightlifting of 20 instances.

#Loading the Required Packages
First, the packages needed are loaded.
```{r}
library(caret)
library(ggplot2)
library(randomForest)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(corrplot)
library(gbm)
```

                    

#Load the data
```{r}
#Download the data
setwd("D:/DOST/")
if(!file.exists("pml-training.csv")){
        fileUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
        download.file(fileUrl,destfile="./pml-training.csv")
}
if(!file.exists("pml-testing.csv")){
        fileUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
        download.file(fileUrl,destfile="./pml-testing.csv")
}
```

We load the datasets.

```{r}
train_in <- read.csv("pml-training.csv")
valid_in <- read.csv("pml-testing.csv")
dim(train_in)
dim(valid_in)
```
As shown below there are 19622 observations and 160 variables in the Training dataset

#Data cleaning
All the variables which contain all NA values are discarded.

```{r}
trainData<- train_in[, colSums(is.na(train_in)) == 0]
validData <- valid_in[, colSums(is.na(valid_in)) == 0]
dim(trainData)
```

```{r}
dim(validData)
```

We now remove the first seven variables as they have little impact on the outcome classe

```{r}
trainData <- trainData[, -c(1:7)]
validData <- validData[, -c(1:7)]
dim(trainData)
```

```{r}
dim(validData)
```

#Preparing the datasets for prediction

Preparing the data for prediction by splitting the training data into 70% as train data and 30% as test data. This splitting will server also to compute the out-of-sample errors.

The test data renamed: valid_in (validate data) will stay as is and will be used later to test the prodction algorithm on the 20 cases.

```{r}
set.seed(1234) 
inTrain <- createDataPartition(trainData$classe, p = 0.7, list = FALSE)
trainData <- trainData[inTrain, ]
testData <- trainData[-inTrain, ]
dim(trainData)
```

```{r}
dim(testData)
```

#Cleaning even further by removing the variables that are near-zero-variance

```{r}
NZV <- nearZeroVar(trainData)
trainData <- trainData[, -NZV]
testData  <- testData[, -NZV]
dim(trainData)
```

```{r}
dim(testData)
```

After this cleaning we are down now to 53 variables

The following correlation plot uses the following parameters (source:CRAN Package 'corrplot') "FPC": the first principal component order. "AOE": the angular order tl.cex Numeric, for the size of text label (variable names) tl.col The color of text label.

```{r}
cor_mat <- cor(trainData[, -53])
corrplot(cor_mat, order = "FPC", method = "color", type = "upper", 
         tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```
In the corrplot graph the correlated predictors (variables ) are those with a dark color intersection.

To obtain the names of the variables we do the following

we use the findCorrelation function to search for highly correlated attributes with a cut off equal to 0.75

```{r}
highlyCorrelated = findCorrelation(cor_mat, cutoff=0.75)
```

We then obtain the names of highly correlated attributes

```{r}
names(trainData)[highlyCorrelated]
```

#Model building
For this project we will use two different algorithms, classification trees and random forests, to predict the outcome.

1. classification trees
2. random forests
3. Generalized Boosted Model

#Prediction with classification trees
We first obtail the model, and then we use the fancyRpartPlot() function to plot the classification tree as a dendogram.

```{r}
set.seed(12345)
decisionTreeMod1 <- rpart(classe ~ ., data=trainData, method="class")
fancyRpartPlot(decisionTreeMod1)
```

We then validate the model "decisionTreeModel" on the testData to find out how well it performs by looking at the accuracy variable.

```{r}
predictTreeMod1 <- predict(decisionTreeMod1, testData, type = "class")
cmtree <- confusionMatrix(predictTreeMod1, testData$classe)
cmtree
```

plot matrix results

```{r}
# plot matrix results
plot(cmtree$table, col = cmtree$byClass, 
     main = paste("Decision Tree - Accuracy =", round(cmtree$overall['Accuracy'], 4)))
```

We see that the accuracy rate of the model is low: 0.6967 and therefore the out-of-sample-error is about 0.3 which is considerable.

#Prediction with Random Forest
We first determine the model

```{r}
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modRF1 <- train(classe ~ ., data=trainData, method="rf", trControl=controlRF)
modRF1$finalModel
```

We then validate the model obtained model "modRF1" on the test data to find out how well it performs by looking at the Accuracy variable

```{r}
predictRF1 <- predict(modRF1, newdata=testData)
cmrf <- confusionMatrix(predictRF1, testData$classe)
cmrf
```

The accuracy rate using the random forest is very high: Accuracy : 1 and therefore the out-of-sample-error is equal to 0***. But it might be due to overfitting.

Let's plot the model
```{r}
plot(modRF1)
```

```{r}
plot(cmrf$table, col = cmrf$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(cmrf$overall['Accuracy'], 4)))
```

#Prediction with Generalized Boosted Regression Models

```{r}
set.seed(12345)
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
modGBM  <- train(classe ~ ., data=trainData, method = "gbm", trControl = controlGBM, verbose = FALSE)
modGBM$finalModel
```

```{r}
# print model summary
print(modGBM)
```

Validate the GBM model and

```{r}
predictGBM <- predict(modGBM, newdata=testData)
cmGBM <- confusionMatrix(predictGBM, testData$classe)
cmGBM
```

The accuracy rate using the random forest is very high: Accuracy : 0.9736 and therefore the *out-of-sample-error is equal to 0.0264**.

#Applying the best model to the validation data

By comparing the accuracy rate values of the three models, it is clear the the 'Random Forest' model is the winner. So will use it on the validation data

```{r}
Results <- predict(modRF1, newdata=validData)
Results
```

